Unsupervised learning of Gaussian mixture models in linear time using a variational Bayesian method

### 2017-05-17 articles/nips-2017/ms.tex
some math

@@

\section{The objective function}

>>>>

\usepackage{amssymb} % for \intercal
\usepackage{bm}      % for bold vectors
\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
\def\veccov{\vect{C}}
\def\vecmean{\vect{\mu}}
\def\vectheta{\vect{\theta}}
\def\weight{w}
\def\weights{\vect{\weight}}
\def\datum{y}
\def\data{\vect{\datum}}
\def\likelihood{\mathcal{L}}
\def\current{^{\mathrm{(K)}}}
\def\future{^{\mathrm{(K} + \Delta\mathrm{K)}}}

@@

\section{The objective function}
\label{sec:objective-function}

\noindent{}The probability density function $f$ for a multivariate normal 
distribution with $D$ dimensions is,

\begin{equation}
  f(\data|\vecmean,\veccov) 
      = \frac{1}{\sqrt{(2\pi)^d|\veccov|}}
        \exp{\left[-\frac{1}{2}(\data - \vecmean)^\intercal\veccov^{-1}(\data - \vecmean)\right]}
\end{equation}

\noindent{}where $\vecmean$ and $\veccov$ is the multivariate mean and 
covariance matrix, and $\data$ are the data.
For a fixed number of $K$ multivariate Gaussian mixtures, the probability
density function is given by:

\begin{equation}
  f(\data|\vectheta) = \sum_{k=1}^{K} \weight_k \, f(\data|\vectheta_k)
\end{equation}

\noindent{}Where $\vectheta_k \equiv \{\vecmean_k, \veccov_k\}$,
$\vectheta \equiv \{\vectheta_1,\dots,\vectheta_K,\weight_1,\dots,\weight_K\}$, 
and $\weight_k$ is the relative weight of the $k$-th mixture such that
$\sum_{k=1}^{K}\weight_k = 1$.

\subsection{Minimum Message Length}
\label{sec:mml}

The Minimum Message Length \citep[MML;][]{Wallace_1968} principle is a formal
description of Occam's razor that is based on information theory.
% Taking Bayes and Shannon.
% TODO
The MML principle states that the best explanation of the data is the model
that describes the data using the smallest amount of information (shortest
message), where information is described as per \citet{Shannon_1948}.

Calculating the optimal lossless message length of a complex model can be
non-trivial or intractable. Here we make use of the \citet{Wallace_1987}
approximation of the message length:

\begin{equation}
\label{eq:message-length}
  I(\vectheta,\data) 
    = \frac{Q}{2}\log\kappa{\left(Q\right)} 
    - \log{\left(\frac{p(\vectheta)}{\sqrt{|\mathcal{F}(\vectheta)|}}\right)}
    - \likelihood\left(\data|\vectheta\right) 
    + \frac{Q}{2} \quad .
\end{equation}

Here $p(\vectheta)$ is the prior, $Q$ is the number of free parameters in the
model, $\kappa\left(Q\right)$ is a function to approximate the lattice
quantisation constant for $Q$ free parameters \citep[e.g.,][]{Conway_1984}, 
and $|\mathcal{F}\left(\vectheta\right)|$ is the determinant of the 
\emph{expected} Fisher information matrix (the second-order partial 
derivatives of the negative log-likelihood function 
$-\likelihood(\data|\vectheta)$).

The MML principle requires that we encode everything required to reconstruct
the message, including our prior beliefs on the model parameters.The message 
for a finite Gaussian mixture must therefore encode:
\begin{enumerate}
  \item The number of Gaussian mixtures, $K$.
  \item The relative weights $\weights$ of the $K - 1$ Gaussian mixtures. Only
        $K - 1$ must be encoded because $\sum_{k=1}^{K}\weight_k = 1$.
  \item The component parameters $\vecmean$ and $\veccov$ for all $K$ Gaussian
        mixtures.
  \item The data, given the model parameters.
  \item The lattice quantisation constant $\kappa(Q)$ for the number of model
        parameters $Q$.
\end{enumerate}

We provide the message length of each component in turn before outlining our
objective function.  First, we assume a prior on $K$ of $p(K) \propto 2^{-K}$
such that the length of the optimal lossless message to encode $K$ is 
$I(K) = -\log{p(K)} = K\log{2} + \textrm{constant}$.  Given a uniform prior
on $\weights$ and the requirement that $\sum_{k=1}^{K}\weight_k = 1$, the 
weights can be treated as parameters of a multinomial distribution, which can
be optimally encoded in a message of length:
\begin{equation}
  I(\weights) 
    = \frac{K - 1}{2}\log{N} 
    - \frac{1}{2}\sum_{k=1}^{K}\log\weight_k 
    - \log{\Gamma{\left(K\right)}} \quad .
\end{equation}

In order to properly encode the mixture parameters $\vecmean$ and $\veccov$
for all $K$ mixtures, we must encode both our prior belief on those parameters,
and the determinant of the expected Fisher information matrix. For the $k$-th
mixture this becomes,

\begin{equation}
  I(\vecmean_k,\veccov_k) = -\log{\left(\frac{p(\vecmean_k,\veccov_k)}{\sqrt{|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|}}\right)}
                          = -\log{p(\vecmean_k,\veccov_k)} + \frac{1}{2}\log{|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|}
\end{equation}

\noindent{}and for all mixtures:

\begin{equation}
  I(\vecmean,\veccov) = -\sum_{k=1}^{K}\log{p(\vecmean_k,\veccov_k)} + \frac{1}{2}\sum_{k=1}^{K}\log{|\mathcal{F}\left(\vecmean_k,\veccov_k\right)|} \quad .
  \label{eq:I_component_params}
\end{equation}

We adopt an improper uniform prior of $\mathcal{U}(\vecmean) = [-\infty, +\infty]$ 
on $\vecmean$, and a conjugate inverted Wishart prior for the covariance matrix
of individual mixtures \citep[e.g., Section 5.2.3. of ][]{Schafer_1997}:
\begin{equation}
  p(\vecmean_k, \veccov_k) \propto |\veccov_k|^{\frac{D+1}{2}} \quad .
  \label{eq:covariance-prior}
\end{equation}

We approximate the determinant of the Fisher information 
$|\mathcal{F}(\vecmean_k, \veccov_k)|$ as the product of 
$|\mathcal{F}\left(\vecmean_k\right)|$ and $|\mathcal{F}\left(\veccov_k\right)|$ 
\citep{Oliver_1996,Roberts_1998}.  Taking the second derivative of the
log-likelihood function $\likelihood\left(\data|\vecmean,\veccov\right)$,
with respect to $\vecmean$ yields,
\begin{equation}
  |\mathcal{F}\left(\vecmean_k\right)| = (N\weight_k)^{D}|\veccov_k|^{-1} \quad .
\end{equation}

We make use of an analytical expression for the determinant of the Fisher
information for a covariance matrix $|\mathcal{F}(\veccov_k)|$ \citep{Dwyer_1967,Magnus_1988,Kasarapu_2015},
\begin{equation}
  |\mathcal{F}\left(\veccov_k\right)| = (N\weight_k)^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)} \quad ,
\end{equation}

\noindent{}giving the following approximation for the determinant of the
expected Fisher information matrix,
\begin{eqnarray}
  |\mathcal{F}(\vecmean_k,\veccov_k)| & \approx & |\mathcal{F}(\vecmean_k)|\cdot|\mathcal{F}(\veccov_k)| \nonumber \\
  |\mathcal{F}(\vecmean_k,\veccov_k)| & \approx & (N\weight_k)^{D}|\veccov_k|^{-1}(N\weight_k)^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)} \nonumber \\
  |\mathcal{F}(\vecmean_k,\veccov_k)| & \approx & (N\weight_k)^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)} \quad .
\end{eqnarray}

\noindent{}which, with Eq. \ref{eq:covariance-prior}, we can substitute into Eq. \ref{eq:I_component_params}:

\begin{eqnarray}
I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\log{|\veccov_k|^{\frac{D + 1}{2}}}
 + \frac{1}{2}\sum_{k=1}^{K}\log{\left[(N\weight_k)^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)}\right]} \nonumber \\
I(\vecmean,\veccov) &=& -\frac{(D + 1)}{2}\sum_{k=1}^{K}\log{|\veccov_k|}
 + \frac{D(D+3)}{4}\sum_{k=1}^{K}\log{\left(N\weight_k\right)} -\frac{KD}{2}\log{2} -\frac{D+2}{2}\sum_{k=1}^{K}\log{|\veccov_k|} \nonumber \\
%I(\vecmean,\veccov) &=& \frac{D(D+3)}{4}\sum_{k=1}^{K}\log{\left(N\weight_k\right)} -\frac{KD}{2}\log{2} -\frac{(D + 1)}{2}\sum_{k=1}^{K}\log{|\veccov_k|}
% +-\frac{D+2}{2}\sum_{k=1}^{K}\log{|\veccov_k|} \nonumber \\
% TODO: UPDATE....
\end{eqnarray}

If we assume that the data have homoskedastic noise properties, then the 
precision of each measurement $\mathcal{E}$ relates the probability of a
datum $\textrm{Pr}(\datum_n)$ can be related to the probability density of
the datum, given the model 
$\textrm{Pr}(\datum_n) = \mathcal{E}^{D}\textrm{Pr}(y_i|\mathcal{M})$.
In the paper we adopt $\mathcal{E} = 0.01$, but we note that the value of
$\mathcal{E}$ has no effect on our inferences.  The message length of encoding
a datum given the model is,
\begin{equation}
  I(\data_n) = -\log{\textrm{Pr}(\data_n)} = -D\log\mathcal{E} - \log\sum_{k=1}^{K}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)
\end{equation}

\noindent{}and for the entire data:
\begin{eqnarray}
  I(\data|\vectheta) &=& -ND\log\mathcal{E} - \sum_{n=1}^{N}\log\sum_{k=1}^{K}w_{k}f_k(\data_n|\vecmean_k,\veccov_k) \quad .
\end{eqnarray}

The final part of the message is to encode the lattice quantisation constant,
which arises from the approximation to the (so-called) strict MML, where
parameters are quantised into intervals in high dimensional space in order to
be losslessly encoded.  We use an approximation of the lattice constant
\citep[see Sections 5.1.12 and 3.3.4 of ][]{Wallace_2005} such that,
\begin{equation}
  \log\kappa(Q) = \frac{\log{Q\pi}}{Q} - \log{2\pi} - 1 \quad ,
\end{equation}

\noindent{}where $Q$ is the total number of free parameters in the model. Each
component requires $D$ parameters to specify $\vecmean$, and a covariance
matrix with off-diagonal terms requires $\frac{1}{2}D(D+1)$ parameters.
Because only $K - 1$ weights need to be encoded, the total number of free
parameters is given by ${Q = K\left[\frac{D(D+3)}{2} + 1\right] - 1}$.

\subsection{Message length for a finite mixture of multivariate Gaussian components}

The message lengths of the optimally encoded components allows us to construct
our full objective function:
\begin{equation}
  I(\vectheta,\data) = I(K) + I(\weights) + \sum_{k=1}^{K}I(\vectheta_k) + I(\data|\vectheta) + \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right)\quad .
\end{equation}

\noindent{}Substituting the expressions outlined in the previous subsection
(except for $Q$, which we leave unexpanded for brevity), we arrive at the full
message length for a finite mixture of $K$ multivariate Gaussian components:
\begin{eqnarray}
  I(\vectheta,\data) &=&
      K\log{2} % I(K)
    + \frac{K - 1}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\log\weight_k - \log{\Gamma(K)} % I(\weights) 
    + \frac{1}{2}\sum_{k=1}^{K}\log{|\veccov_k|} \nonumber \\ % 
   &+&\frac{K}{2}\left[\frac{D(D+3)}{2}\log{N} - D\log{2}\right] % I(\vectheta_k) = I(\vecmean_k,\veccov_k)
    - ND\log\mathcal{E} - \sum_{n=1}^{N}\log\sum_{k=1}^{K}w_{k}f_k(\data_n|\vecmean_k,\veccov_k) \nonumber \\ % I(\data|\vectheta)
   &+& \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) \quad . % lattice
    \label{eq:objective-function}
\end{eqnarray}

\subsection{Predicting the message length of uncomputed mixtures}

Consider that the number of true components $K_{true}$ is unknown, and that we
have an optimal mixture of $K$ components in the sense that the model
parameters $\vectheta$ minimise the total message length in Eq.
\ref{eq:objective-function}.  We are interested to know whether the total
message length will decrease if we add or remove components.  For
example, our current state may be an optimized mixture with $K=2$ and we want
to know if an optimal mixture with $K=3$ will result in a shorter message
length.  By denoting $I\current$ as the message length of the current (optimal) 
mixture and $I\future$ as mixtures that have yet to be computed, we
want to know if $I\future - I\current < 0$, or rather we seek to calculate 
some probability $\textrm{Pr}(I\future - I\current < 0)$ before having to 
optimize the future mixture $I\future$.  
Taking Eq. \ref{eq:objective-function}, we can derive an expression for the 
difference in message lengths between future mixtures and the current state:
\begin{eqnarray}
  I\future - I\current &=& \Delta{K}\left[
  \left(1 - \frac{D}{2}\right)\log{2} + \frac{1}{4}\left(D(D+3)+2\right)\log\frac{N}{2\pi}\right] \nonumber \\
  &-& \log\Gamma(K+\Delta{K}) + \log\Gamma(K) + \frac{1}{2}\left(\log{Q\future} - \log{Q\current}\right) \nonumber \\
  &+& \frac{1}{2}\left(\frac{D(D+3)}{2} - 1\right)\left(\sum_{k=1}^{K+\Delta{K}}\log\weight_k\future - \sum_{k=1}^{K}\weight_k\current\right) \nonumber \\
  &-& \sum_{n=1}^{N}\log\sum_{k=1}^{K+\Delta{K}}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)\future + \sum_{n=1}^{N}\log\sum_{k=1}^{K}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)\current \nonumber \\
  &+& \frac{D + 2}{2}\left(\sum_{k=1}^{K}\log|\veccov_k|\current - \sum_{k=1}^{K + \Delta{K}}\log|\veccov_k|\future\right) \quad .
\end{eqnarray}
